Composition-climate paper

### 2016-11-27 Response_to_reviewers.Rmd

> Another issue is lack of discussion about the time period when the forests sampled by FIA got established. It is typical in SDM studies to test the strength of models from different time periods in tree SDM work and select the time period that is most strongly associated with species presence.  In the western US, we found 1950-1980 to produce stronger models than 1980-2010.

We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks, we believe that our contribution further underscores this point.

The normals for the region, beginning with a normal ending 1980 and continuing to a normal ending 2014, give a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.  These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively, but, as pointed out in Baker *et al*. (2016; doi:10.1111/gcb.13273) uncertainty in the underlying estimates are often overlooked.  Given the variability within windows, and the relatively young age of most stands in WI, MN and MI (FIA data shows mean ages of 52, 56 and 49 years respectively), we believe that the choice of normal is acceptable.  Having said that, we examined the importance using both shifts in PLS era, and shifts in the modern climate normal.  The largest effect is in $t_{max}$, since the most recent normals all show a significant increase in $t_{max}$.

We have added the sentence: *"The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study."*

@@

*Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood*

We believe that the issue of stability or extending the dist-clim relationship is a matter of opinion.  The paper shows that, for many taxa, land use following EuroAmerican settlement in the north-central US limits the representitiveness of modern inventory-based distribution models.  We do believe that these historical datases can be used to extend distribution-climate relationships.  The exlicit tests in this paper focus on shifts in climate-distribution relationships and the attribution for these shifts, thus, the stabilty of the relationships.

> 3) One of the objectives in this manuscript was to investigate the effects of climate change, land use change, and pathogen outbreaks on distribution changes (L125-126). However, the pathogen outbreaks were nearly ignored in the title and methods, results, and discussion sections.

We agree.  The impact of pathogen outbreaks are significant for *Ulmus*, and (outside the current range) for *Castanea*, however, given the current data sources, they are difficult to properly assess.  We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions in the Discussion and in the Introduction.

@@

> Most importantly, the way that the authors included the effects of land use change in data analyses (let’s forget pathogen outbreaks for now) was not justified. Thus, I was not convinced the data analyses were robust to answer the question about the effects of climate change, land use change, and pathogen outbreaks on distribution changes. Although the authors argued the land use effects were represented as the difference between VHCH, VMCH. Why?

Previous authors have clearly indicated that EuroAmerican settlement in the north-central United States can be almost entirely attributed to human agency.  In responding to Reviewer 2, we make the same point, and make this point again below.  EuroAmerican Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture.  Thus, VH, from the PLSS forest survey (ignoring climate data for now), represents forest structure and distribution in the absence of settlement impacts.  VM represents a modern landscape, sampled over the last decade as part of the FIA program.  Distribution loss is almost entirely attributable to anthropogenic agency, either loss of forested land (oak-hickory forest lost 5x10^6^ ha of coverage since settlement due to agricultural conversion, as pointed out in the response to Reviewer 2), thus we feel confident in ascribing change from VC -> VM to land use.  We believe that changes in the introduction and discussion, strengthening the argument for widespread vegetation change help resolve this issue.

### 2016-12-02 Composition_Ecology.Rmd
I had to soft-reset a bunch of changes to unde the accidental inclusion of two large RDS files.
Changes include all revisions up to the last pass-off to @iceageecologist & include code changes to improve the modularity of the code.

# Effect of historical land-use and climate change on tree-climate relationships in the north-central United States

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. These distributions of species in geographic and environmental space, combined with future climate scenarios, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, the predictive capacity of ecological models is being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for processes operating at large temporal and spatial scales [@moorcroft2006close;@matthes2016benchmarking].

@@

Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure occurring across the species' environmental gradient. Predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions. Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. Land use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.).

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change and land use, and we diagnose the relative influence of historical climate and land use changes on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect. FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset]. PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].

Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.  In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.  For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.

@@

We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] to generate historical and modern climate data. The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.  We resample estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) to the 8 x 8km grid used for PLS and FIA data. These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].  Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.

Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).  Baker et al. [-@baker1985minnesota] show long term increases in mean annual temperature at Ft. Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas [@burnette2010daily] show increasing temperatures in all seasons, with faster increases during winter (0.10^o^C decade^-1^) than during summer (0.07^o^C decade^-1^).  This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.  The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study.

@@

Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) of tree species presence ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions). Hellinger distance for two discrete probability distributions is defined as:

@@

where $P$ and $Q$ are the distributions, with a common discrete index $i$.

The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). These conditions are observed or recorded in data.  The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.  The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.  We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.

We develop an attribution index, ($\delta_{d} = d_{c} - d_{v}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.

To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If either or both changes are not significant, then the change is defined as confounding.

@@

All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.

@@

#  This takes a bit of time.
loss_melt <- loss_plot()

@@

boxplot_clim <- clim_boxplots(clim_table = allveg,
                              filename = "boxplot_corrected_all")

>>>>

# Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. These distributions of species in geographic and environmental space, combined with future climate scenarios, are used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for processes operating at long time scales [@moorcroft2006close;@matthes2016benchmarking].

@@

Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient. For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.  These confounding effects may hinder detection of climatic forcing of historical species distributions. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions. Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. Land use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review these methods. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees. Goring et al. [-@goring2016preset] cleaned and aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA plots are sited within forested or partially forested landscapes [@woudenberg2010forest]. while the PLS sampling is uniform. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the assignments of 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.  Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.  The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.  Both the FIA and PLS datasets both have good coverage in the upper Midwest, with   8km grid cells with at least one FIA plot and XXX  grid cells with at least one PLS point.  In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.  We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA data (Supplementary Table  2). For all species, the range within the upper Midwest incompletly samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.

@@

We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] for the historical and modern climate data. The PRISM climate variables are available at 800m resolution and are interpolated from station locations using elevation, aspect and other data.  We resample estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) to the 8 x 8km grid used for PLS and FIA data. These climate variables represent three major climatic controls on plant distributions: summer warmth, winter minimum temperatures, and moisture availability; these or similar bioclimatic variables are widely used in plant distributional modeling [@woodward1987climate;@iverson1998predicting;@maiorano2013building].

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 and 1985 - 2014.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].

The use of alternative modern climate normals reflects two reasonable choices for whether to pick climate normals corresponding to time of sampling versus time of establishment.  The 1985-2014 normal encompasses the period of sampling (1998 - 2015) for the FIA data.  The 1950  - 1979 normal encompases the time of establishment of FIA trees, which have an average age of 50 years in this region based on the raw stand data.  We report results for the 1985 - 2014 normals, with the 1950-1980 results show in in Supplmentary information.

@@

Distributions of tree genera within climate space are shown using box plots and described as univariate probability density functions (PDFs) of tree species presence in vegetation ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.  Differences among univariate PDFs are quantified for each climate variable using Hellinger distances [@matthes2016benchmarking] among the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions). The Hellinger distance for two discrete probability distributions is:

@@

where $P$ and $Q$ are the distributions, with a common index $i$.

The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.  The effect of climate change on shifting PDFs is $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The effect of vegetation change on shifting PDFs is similarly described as $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a). Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with EuroAmerican settlement.

We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.

To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If climate change is significant, but land use is not significant, then the change is defined as confounding.  <!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed. -->

@@

All analyses use R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.

@@

#  This takes a bit of time.
if ('loss_melt.rds' %in% list.files('data/output/')) {
  loss_melt <- readRDS('data/output/loss_melt.rds')
} else {
  loss_melt <- loss_plot()
  saveRDS(loss_melt, file = 'data/output/loss_melt.rds')
}

@@

boxplot_allveg <- clim_boxplots(clim_table = allveg,
                              filename = "boxplot_corrected_all")

boxplot_plss <- clim_boxplots(clim_table = newveg,
                              filename = "boxplot_corrected_pls")
